rang: Reconstructing reproducible R computational environments

### 2023-05-16 paper/paper.qmd
First pass

---
title: 'rang: Reconstructing reproducible R computational environments'
tags:
  - R
  - reproducibility
  - docker
authors:
  - name: "Chung-hong Chan"
    orcid: 0000-0002-6232-7530
    affiliation: GESIS Leibniz-Institut für Sozialwissenschaften
    email: chung-hong.chan@gesis.org
    corresponding: true
  - name: "David Schoch"
    orcid: 0000-0003-2952-4812
    affiliation: GESIS Leibniz-Institut für Sozialwissenschaften
abstract: |
  A complete declarative description of the computational environment is often missing when researchers share their materials. Without such description, software obsolescence and missing system components can jeopardize computational reproducibility in the future, even when data and computer code are available. The R package rang is a complete solution for generating the declarative description for other researchers to automatically reconstruct the computational environment at a specific time point. The reconstruction process, based on Docker, has been tested for R code as old as 2001. The declarative description generated by rang satisfies the definition of a reproducible research compendium and can be shared as such. In this contribution, we show how rang can be used to make otherwise unexecutable code, spanning from fields such as computational social science and bioinformatics, executable again. We also provide instructions on how to use rang to construct reproducible and shareable research compendia of current research. The package is currently available from CRAN (https://cran.r-project.org/web/packages/rang/index.html) and GitHub (https://github.com/chainsawriot/rang).
keywords:
  - R
  - reproducibility
  - docker
year: 2023
bibliography: paper.bib
format:
  pdf: default
  plos-pdf:
    keep-tex: true
    journal:
      id: plosone
---

@@

Other than bad programming practices [@trisovic:2022], the main computing barrier to computational reproducibility is the failure to reconstruct the computational environment like the one used by the original researchers. This task looks trivially simple. But as computer science research has shown, this task is incredibly complex [@abate:2015:MCR;@dolstra:2010:N]. For a usual scripting language such as R [^r], that pertains to four aspects: A) operating system, B) system components such as `libxml2`, C) the exact R version, and D) what and which version of the installed R packages. We will call them Component A, B, C, D in the following sections. Any change in these four components can possibly affect the execution of any shared computer code. For example, the lack of the system component `libxml2` can impact whether the R package `xml2` can be installed on a Linux system. If the shared computer code requires the R package `xml2`, then the whole execution fails.

@@

In reality, the impact of Component A is relatively weak as mainstream, open source programming languages and their software libraries are usually cross platform. In modern computational research, Linux is the de-facto operating system in high performance computing environments (e.g. Slurm). Instead, the impact of Components B, C, and D is much higher. Component D is the most volatile among them all as there are many possible combinations of R packages and versions. Software updates with breaking changes (even in a dependency) might render existing shared code using those changed features not executable or not producing the same result anymore. Also, software obsolescence is commonplace, especially since academic software is often not well maintained due to lack of incentives [@merow:2023:B].

@@

To build a container, one needs to write a plain text declarative description of the required computational environment. Inside this declarative description, it should pin down all four Components mentioned above. For Docker, it is in the form of a plain text file called `Dockerfile`. This `Dockerfile` is then used as the recipe to build a Docker image, where the four Components are assembled. Then, one can launch a container with the built Docker image.

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of `Dockerfile` is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a `Dockerfile` [@kim:2018:E]. But there are many tools available to automate the process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can pin down and restore all four components without the reliance on any commercial service <!-- such as MRAN -->.

@@

`renv` [@renvrpkg] (and its derivatives such as `jetpack` and its predecessor `packrat`) takes a similar approach to Python's `virtualenv` and Ruby's `Gem` to pin down the exact version of R packages using a "lock file". Other solutions such as `checkpoint` [@checkpointrpkg] depend on the availability of The Microsoft R Application Network (MRAN, a time-stamped daily backup of CRAN), which will be shut down on July 1st, 2023. `groundhog` [@groundhogrpkg] used to depend on MRAN but has a plan to switch to their home-grown R package repository. These solution can effectively pin down Component C and D. But they can only restore component D. Also, for solutions depending on MRAN, there is a limit on how far back this reproducibility can go, since MRAN can only go back as far as September 17, 2014. Additionally, it only covers CRAN packages.

`containerit` [@nuest:2019] takes the current state of the computational environment and documents it as a `Dockerfile`. `containerit` makes the assumption that Component A has a weak influence on computational reproducibility and therefore defaults to Linux-based Rocker base images [@boettiger:2017:IR]. In this way, it fixes Component A. But `containerit` does not pin down the exact version of R packages. Therefore, it can pin down components A, B, C, but only a part of component D. `dockta` [^dockta] is another containerization solution that can potentially pin down all components due to the fact that MRAN is used. But it also suffers from the same limitations mentioned above.

@@

There are two important functions of `rang`: `resolve()` and `dockerize()`.

@@

All data and computer code are available from GitHub with the last commit on 2019-01-17 [^hilgard]. The R code contains a list of R packages used in the project as `library()` statements, including an R package on GitHub that is written by the main author of that paper. However, we identified one package (`compute.es`) that was not written in those `library()` statements but used with the namespace operator, i.e. `compute.es::tes()`. This undocumented package can be detected by `renv::dependencies()`, which is the provider of the scanning function of `rang`.

@@

The R package `maxent` introduces a machine learning algorithm with a small memory footprint and was available on CRAN until 2019. A software paper was published by the original authors in 2012 [@jurka:2012]. The R package was also used in some subsequent automated content analytic papers [e.g. @loercher:2017:D]. Despite the covert editing of the package by a staffer of CRAN [^evidence], the package was removed from CRAN in 2019 [^checkerror]. We attempted to install the second last (the original submitted version) and last (with covert editing) versions of `maxent` on R 4.2.2. Both of them didn't work.

@@

```r
require(ptproc)

@@

```bash
Rscript -e "require(rang); dockerize(resolve('ptproc', '2004-07-01'),
'pengdocker', cache = TRUE)"
docker build -t pengimg ./pengdocker
## launching a container in daemon mode -d
docker run -d --rm --name "pengcontainer" -ti pengimg
docker cp peng.R pengcontainer:/peng.R
docker exec pengcontainer R CMD BATCH peng.R
docker exec pengcontainer cat peng.Rout
docker cp pengcontainer:/peng.Rout peng.Rout
docker stop pengcontainer
```

@@

```sh
Rscript -e "require(rang); dockerize(resolve('Sushi', '2014-06-05'),
'sushidocker', no_rocker = TRUE, cache = TRUE)"
docker build -t sushiimg ./sushidocker
docker run -d --rm --name "sushicontainer" -ti sushiimg
docker cp PaperFigure.R sushicontainer:/PaperFigure.R
docker exec sushicontainer mkdir vignettes
docker exec sushicontainer R CMD BATCH PaperFigure.R
docker cp sushicontainer:/vignettes/Figure_1.pdf sushi_figure1.pdf
docker stop sushicontainer
```

@@

To demonstrate this point, we took the recent paper by Oser et al. [@oser:2022:HPE]. This paper was selected because 1) the paper was published in *Political Communication*, a high impact journal that awards Open Science Badges; 2) shared data and R code are available; and most importantly, 3) the shared R code is well-written. In the repository of this paper, we based on the materials shared by Oser et al. [@oser:2022:HPE] and prepared a research compendium that should have long-term computational reproducibility. The research compendium is similar to the Executable Compendium suggested by the Turing way.

@@

```r
require(rang)
## meta-analysis is the directory of all shared materials
cran_pkgs <- as_pkgrefs("meta-analysis")

@@

In this executable compendium, only the first four elements are essential. The directory `oserdocker` (116 MB) contains cached R packages, a `Dockerfile`, and a verbatim copy of the directory `meta-analysis/` to be transferred into the Docker image. That can be regenerated by running `make resolve`. However, having this directory preserved insures against the situations that some R packages used in the project were no longer available or any of the information providers used by `rang` for resolving the dependency relationships were not available. (Or in the rare circumstance of `rang` is no longer available.)

@@

There are still many features that we did not mention in this paper. `rang` is built with interoperability in mind. As of writing, `rang` is interoperable with existing R packages such as `renv` and R built-in `sessionInfo()`. Also, the `rang` object can be used for network analysis with R packages such as `igraph`.

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs. The (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. There are also other limitations. In the Vignette of `rang` ([https://cran.r-project.org/web/packages/rang/vignettes/faq.html](https://cran.r-project.org/web/packages/rang/vignettes/faq.html)), we list all of these limitations as well as possible mitigation.

>>>>

---
title: 'rang: Reconstructing reproducible R computational environments'
tags:
  - R
  - reproducibility
  - docker
authors:
  - name: "Chung-hong Chan"
    orcid: 0000-0002-6232-7530
    affiliation: GESIS Leibniz-Institut für Sozialwissenschaften
    email: chung-hong.chan@gesis.org
    corresponding: true
  - name: "David Schoch"
    orcid: 0000-0003-2952-4812
    affiliation: GESIS Leibniz-Institut für Sozialwissenschaften
abstract: |
  A complete declarative description of the computational environment is usually missing when researchers share their materials. Without such description, software obsolescence and missing system components can jeopardize computational reproducibility in the future, even when data and computer code are available. The R package rang is a complete solution for generating the declarative description for other researchers to automatically reconstruct the computational environment at a specific time point. The reconstruction process, based on Docker, has been tested for R code as old as 2001. The declarative description generated by rang satisfies the definition of a reproducible research compendium and can be shared as such. In this contribution, we show how rang can be used to make otherwise unexecutable code, spanning fields such as computational social science and bioinformatics, executable again. We also provide instructions on how to use rang to construct reproducible and shareable research compendia of current research. The package is currently available from CRAN (https://cran.r-project.org/web/packages/rang/index.html) and GitHub (https://github.com/chainsawriot/rang).
keywords:
  - R
  - reproducibility
  - docker
year: 2023
bibliography: paper.bib
format:
  pdf: default
  plos-pdf:
    keep-tex: true
    journal:
      id: plosone
---

@@

Other than bad programming practices [@trisovic:2022], the main computing barrier to computational reproducibility is the failure to reconstruct the computational environment like the one used by the original researchers. This task looks trivially simple. But as computer science research has shown, this task is incredibly complex [@abate:2015:MCR;@dolstra:2010:N]. For a usual scripting language such as R [@rcore] [^r], that pertains to four aspects: A) operating system, B) system components such as `libxml2`, C) the exact R version, and D) what and which version of the installed R packages. We will call them Component A, B, C, D in the following sections. Any change in these four components can possibly affect the execution of any shared computer code. For example, the lack of the system component `libxml2` can impact whether the R package `xml2` can be installed on a Linux system. If the shared computer code requires the R package `xml2`, then the whole execution fails.

@@

In reality, the impact of Component A is relatively weak as mainstream, open source programming languages and their software libraries are usually cross platform. In modern computational research, Linux is the de-facto operating system in High Performance Computing environments. Instead, the impact of Components B, C, and D is much higher. Component D is the most volatile among them all as there are many possible combinations of R packages and versions. Software updates with breaking changes (even in a dependency) might render existing shared code using those changed features not executable or not producing the same result anymore. Also, software obsolescence is commonplace, especially since academic software is often not well maintained due to lack of incentives [@merow:2023:B].

@@

To build a container, one needs to write a plain text declarative description of the required computational environment. Inside this declarative description, it should unambiguously specify all four Components mentioned above. For Docker, it is in the form of a plain text file called `Dockerfile`. This `Dockerfile` is then used as the recipe to build a Docker image, where the four Components are assembled. Then, one can launch a container with the built Docker image.

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of `Dockerfile` is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a `Dockerfile` [@kim:2018:E]. But there are many tools available to automate the process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can specify and restore all four components without the reliance on any commercial service <!-- such as MRAN -->.

@@

`renv` [@renvrpkg] (and its derivatives such as `jetpack` and its predecessor `packrat`) takes a similar approach to Python's `virtualenv` and Ruby's `Gem` to unambiguously specify the exact version of R packages using a "lock file". Other solutions such as `checkpoint` [@checkpointrpkg] depend on the availability of The Microsoft R Application Network (MRAN, a time-stamped daily backup of CRAN), which will be shut down on July 1, 2023. `groundhog` [@groundhogrpkg] used to depend on MRAN but has a plan to switch to their home-grown R package repository. These solution can effectively specify Component C and D. But they can only restore component D. Also, for solutions depending on MRAN, there is a limit on how far back this reproducibility can go, since MRAN can only go back as far as September 17, 2014. Additionally, it only covers CRAN packages.

`containerit` [@nuest:2019] takes the current state of the computational environment and documents it as a `Dockerfile`. `containerit` makes the assumption that Component A has a weak influence on computational reproducibility and therefore defaults to Linux-based Rocker base images [@boettiger:2017:IR]. In this way, it fixes Component A. But `containerit` does not specify the exact version of R packages. Therefore, it can specify components A, B, C, but only a part of component D. `dockta` [^dockta] is another containerization solution that can potentially specify all components due to the fact that MRAN is used. But it also suffers from the same limitations of MRAN mentioned above, i.e. shutting down on July 1, 2023, going back as far as September 17, 2014, and covering only CRAN.

@@

There are two important functions of `rang` version 0.2.0: `resolve()` and `dockerize()`.

@@

All data and computer code are available from GitHub with the last commit on 2019-01-17 [^hilgard]. The R code contains a list of R packages used in the project as `library()` statements, including an R package on GitHub that is written by the main author of that paper. However, we identified one package (`compute.es`) that was not written in those `library()` statements but used with the namespace operator, i.e. `compute.es::tes()`. This undocumented package can be detected by `renv::dependencies()`, of which `as_pkgrefs()` is a wrapper.

@@

The R package `maxent` introduces a machine learning algorithm with a small memory footprint and was available on CRAN until 2019. A software paper was published by the original authors in 2012 [@jurka:2012]. The R package was also used in some subsequent automated content analytic papers [e.g. @loercher:2017:D]. Despite the editing of the package by a staffer of CRAN [^evidence], the package was removed from CRAN in 2019 [^checkerror]. We attempted to install the second last (the original submitted version) and last (with editing) versions of `maxent` on R 4.2.2. Neither worked.

@@

```r
library(ptproc)

@@

```bash
Rscript -e "library(rang); dockerize(resolve('ptproc', '2004-07-01'),
'pengdocker', cache = TRUE)"
docker build -t pengimg ./pengdocker
## launching a container in daemon mode -d
docker run -d --rm --name "pengcontainer" -ti pengimg
docker cp peng.R pengcontainer:/peng.R
docker exec pengcontainer R CMD BATCH peng.R
docker exec pengcontainer cat peng.Rout
docker cp pengcontainer:/peng.Rout peng.Rout
docker stop pengcontainer
```

@@

```sh
Rscript -e "library(rang); dockerize(resolve('Sushi', '2014-06-05'),
'sushidocker', no_rocker = TRUE, cache = TRUE)"
docker build -t sushiimg ./sushidocker
docker run -d --rm --name "sushicontainer" -ti sushiimg
docker cp PaperFigure.R sushicontainer:/PaperFigure.R
docker exec sushicontainer mkdir vignettes
docker exec sushicontainer R CMD BATCH PaperFigure.R
docker cp sushicontainer:/vignettes/Figure_1.pdf sushi_figure1.pdf
docker stop sushicontainer
```

@@

To demonstrate this point, we took the recent paper by Oser et al. [@oser:2022:HPE]. This paper was selected because 1) the paper was published in *Political Communication*, a high impact journal that awards Open Science Badges; 2) shared data and R code are available; and most importantly, 3) the shared R code is well-written. In the repository of this paper, we based on the materials shared by Oser et al. [@oser:2022:HPE] and prepared a research compendium that should have long-term computational reproducibility. The research compendium is similar to the Executable Compendium suggested by the Turing way [@community:2019:TW].

@@

```r
library(rang)
## meta-analysis is the directory of all shared materials
cran_pkgs <- as_pkgrefs("meta-analysis")

@@

In this executable compendium, only the first four elements are essential. The directory `oserdocker` (116 MB) contains cached R packages, a `Dockerfile`, and a verbatim copy of the directory `meta-analysis/` to be transferred into the Docker image. That can be regenerated by running `make resolve`. However, preserving this directory insures against the situations where (1) some R packages used in the project are no longer available, (2) any of the information providers used by `rang` for resolving the dependency relationships are not available, or (3) in the rare circumstance of `rang` is no longer available.

@@

There are still many features that we did not mention in this paper (see Table 1). `rang` is built with interoperability in mind. As of writing, `rang` is interoperable with existing R packages such as `renv` and R built-in `sessionInfo()`. Also, the `rang` object can be used for network analysis with R packages such as `igraph`.

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs and cannot solve all issues. Specifically, `rang` can restore a specific computational environment but there might be potential barriers to reproducibility related to external dependencies that cannot be containerized. A common external dependency is an external data source connected via a web application programming interface (API). Some examples are the so-called "API packages" such as `STRINGdb` (an R interface to the STRING protein-protein database), `rtoot` (an R interface to the Mastodon API), and `googleLanguageR` (an R interface to the Google Translation API). Although these API packages can be containerized, the external data sources cannot. Therefore, `rang` is not a solution to issues related to external dependencies. For data collection via these API packages, we recommend caching the collected data and sharing the cached data. For computation via API packages (e.g. Google Translate), we recommend using alternative methods that do not require external dependencies (e.g. [@chan:2020:REC]).

Moreover, the (re)construction process based on `rang` takes notably more time than other solutions because all packages are compiled from source. `rang` trades computational efficiency of this often one-off (re)constructing process for correctness, backward compatibility and independence from any commercial backups of software repositories such as MRAN. In the Vignette of `rang` ([https://cran.r-project.org/web/packages/rang/vignettes/faq.html](https://cran.r-project.org/web/packages/rang/vignettes/faq.html)), we list all limitations: (1) `rang` does not support R < 1.3.1, (2) `rang` can generate Ubuntu/Debian-based Docker images and therefore non-Linux features (e.g. WinBUGS) are not supported, (3) `dockerize()` cannot cache OS-level packages (`deb` packages), (4) R packages with changing system requirements might be incorrectly queried, (5) caching of R packages and R source must be enforced when the R version is too old, (6) R packages on CRAN, Bioconductor, and Github, as well as Rocker images, might not be available in the future. All caveats listed have no solution, except point (6) can be mitigated by backing up the generated Docker image (see the section on research compendia).

| Function           | Purpose                                                                              |
|--------------------|--------------------------------------------------------------------------------------|
| `as_pkgrefs`       | Convert data structures (e.g. `sessionInfo`) into package references for `resolve()` |
| `convert_edgelist` | Convert the resolved results to edgelist                                             |
| `dockerize`        | Dockerize the resolved result                                                        |
| `export_rang`      | Export the resolved result as installation script                                    |
| `export_renv`      | Export the resolved result as renv lockfile                                          |
| `query_sysreqs`    | Query for system requirements                                                        |
| `resolve`          | Resolve dependencies of R packages                                                   |

: List of all functions in rang version 0.2.0
